---
title: "Topic 5: Data Cleaning and filtering"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
In this topic, you will learn about :

- Keep and remove variables
- Filtering data and variables

# {.tabset}

## Keep and remove variables

### Keeping and Removing Variables in R Programming

In R, you can selectively keep or remove variables (columns) from a data frame using various functions and techniques. This allows you to manipulate and subset data to work with only the variables you need for analysis or visualization.

**Keeping Variables:**

Using $ Operator: You can keep specific variables by using the $ operator to select them from the data frame.

Example: Keeping Variables using $ Operator

```{r}
# Sample data frame
data <- data.frame(
  ID = 1:5,
  Age = c(25, 30, 22, 28, 35),
  Height = c(170, 180, 165, 175, 190),
  Weight = c(70, 80, 60, 75, 85)
)

# Keep only the Age and Height variables
selected_data <- data[c("Age", "Height")]

```

Using Square Brackets: You can use square brackets with column names to keep specific variables from the data frame.

Example: Keeping Variables using Square Brackets

```{r}
# Sample data frame
data <- data.frame(
  ID = 1:5,
  Age = c(25, 30, 22, 28, 35),
  Height = c(170, 180, 165, 175, 190),
  Weight = c(70, 80, 60, 75, 85)
)

# Keep only the Age and Weight variables
selected_data <- data[, c("Age", "Weight")]

```

**Removing Variables:**

Using $ Operator or Square Brackets with Negation: You can remove specific variables by using the $ operator or square brackets with a negation sign (-) in front of the column names.

Example: Removing Variables using $ Operator and Square Brackets with Negation

```{r}
# Sample data frame
data <- data.frame(
  ID = 1:5,
  Age = c(25, 30, 22, 28, 35),
  Height = c(170, 180, 165, 175, 190),
  Weight = c(70, 80, 60, 75, 85)
)

# Remove the ID variable
selected_data <- data[c("Age", "Height", "Weight")]

# Alternative method: Remove the ID variable
selected_data <- data[-1]

```

Using select() from dplyr: If you are using the dplyr package, you can use the select() function to remove variables.

Example: Removing Variables using select() from dplyr

```{r}
# Load dplyr library
library(dplyr)

# Sample data frame
data <- data.frame(
  ID = 1:5,
  Age = c(25, 30, 22, 28, 35),
  Height = c(170, 180, 165, 175, 190),
  Weight = c(70, 80, 60, 75, 85)
)

# Remove the ID variable
selected_data <- select(data, -ID)

```

Summary:

You can keep specific variables using the $ operator, square brackets, or the select() function from dplyr.
To remove variables, use the negation sign with the $ operator, square brackets, or the select() function from dplyr.
Choose the method that best suits your data manipulation needs and the packages you are using.

## Filtering data and variables

### Filtering Data and Variables in R Programming


Filtering data and variables in R involves selecting specific rows or columns from a data frame based on certain conditions or criteria. This process allows you to extract subsets of the data for further analysis or visualization.

**Filtering Data (Rows):**

Using Logical Indexing: You can use logical indexing to filter rows based on specific conditions.

Example: Filtering Data using Logical Indexing

```{r}
# Sample data frame
data <- data.frame(
  ID = 1:5,
  Age = c(25, 30, 22, 28, 35),
  Gender = c("Male", "Female", "Male", "Male", "Female")
)

# Filter rows where Age is greater than 25
filtered_data <- data[data$Age > 25, ]

```

**Filtering Variables (Columns):**

Using $ Operator or Square Brackets: You can use the $ operator or square brackets to select specific columns from a data frame.

Example: Filtering Variables using $ Operator and Square Brackets

```{r}
# Sample data frame
data <- data.frame(
  ID = 1:5,
  Age = c(25, 30, 22, 28, 35),
  Gender = c("Male", "Female", "Male", "Male", "Female")
)

# Select only the Age and Gender columns
selected_variables <- data[c("Age", "Gender")]

```

Using select() from dplyr: If you are using the dplyr package, you can use the select() function to filter columns.

Example: Filtering Variables using select() from dplyr

```{r}
# Load dplyr library
library(dplyr)

# Sample data frame
data <- data.frame(
  ID = 1:5,
  Age = c(25, 30, 22, 28, 35),
  Gender = c("Male", "Female", "Male", "Male", "Female")
)

# Select only the Age and Gender columns
selected_variables <- select(data, Age, Gender)

```

Summary:

Filtering data (rows) involves selecting specific rows based on conditions using logical indexing or the filter() function from dplyr.
Filtering variables (columns) involves selecting specific columns using the $ operator, square brackets, or the select() function from dplyr.
Choose the appropriate method based on your data manipulation needs and whether you are using the dplyr package for data manipulation.

