---
title: 'Topic 4: Programming statistical graphics'
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: united
    highlight: tango
    df_print: paged
    includes:
      before_body: header.html
      after_body: footer.html
      in_header: GA_script.html

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



In this topic in week 7, you will learn to plot using ggplot2 package (graphics) focus on: 

- Histogram
- Scatter diagram
- Bar chart 
- Box plot
- Line chart


# {.tabset}

## Histogram

**Histogram using ggplot2 in R Programming**

ggplot2 is a powerful and popular data visualization package in R that provides a flexible and elegant way to create high-quality graphics. To create a histogram using ggplot2, you need to follow a specific workflow using the **ggplot()** function along with the **geom_histogram()** function.

**Step-by-Step Workflow:**

Load the necessary library: You need to load the **ggplot2** library before using it.

```{r}
# Load ggplot2 library
library(ggplot2)

```

1. Prepare the data: Ensure you have the data in a data frame or tibble format with the variable you want to plot.
2. Create the ggplot object: Use the **ggplot()** function to specify the data and aesthetic mappings for the plot.
3. Add the geometry: Use the **geom_histogram()** function to create the histogram.
4. Customize the plot (optional): You can customize the appearance of the plot using various ggplot2 functions like **labs(), theme(), scale_x_continuous(), scale_y_continuous(),** etc.

**Example: Creating a Histogram with ggplot2**

```{r}
# Sample data
data <- data.frame(values = c(10, 15, 12, 18, 20, 25, 22, 17, 14, 30))

# Create the ggplot object and add the histogram geometry
ggplot(data, aes(x = values)) +
  geom_histogram(fill = "steelblue", color = "black", bins = 5) +
  labs(title = "Histogram Example", x = "Values", y = "Frequency")

```

In this example, we have a sample data frame named data with a single column values. We created a histogram using **ggplot2** by specifying the data and aesthetic mappings in the **ggplot()** function and adding the histogram geometry with **geom_histogram()**. We customized the fill color, border color, and number of bins in the histogram. We also added a title and axis labels using **labs()**.

Remember to replace data and values with your actual data frame and variable name in the **ggplot2** code. You can further customize the appearance of the plot using additional **ggplot2** functions to meet your specific visualization needs. **ggplot2** provides a wide range of options for creating beautiful and informative histograms.


## Scatter Plot

**Scatterplot using ggplot2 in R Programming**

ggplot2 is a powerful data visualization package in R that offers a flexible and intuitive way to create high-quality graphics. To create a scatterplot using ggplot2, you need to follow a specific workflow using the ggplot() function along with the geom_point() function.

**Step-by-Step Workflow:**

Load the necessary library: You need to load the **ggplot2** library before using it.

```{r}
# Load ggplot2 library
library(ggplot2)

```

1. Prepare the data: Ensure you have the data in a data frame or tibble format with the variables you want to plot.
2. Create the ggplot object: Use the **ggplot()** function to specify the data and aesthetic mappings for the plot.
3. Add the geometry: Use the **geom_point()** function to create the scatterplot.
4. Customize the plot (optional): You can customize the appearance of the plot using various **ggplot2** functions like **labs(), theme(), scale_x_continuous(), scale_y_continuous(),** etc.

**Example: Creating a Scatterplot with ggplot2**

```{r}
# Sample data
data <- data.frame(x_values = c(1, 2, 3, 4, 5, 6),
                   y_values = c(10, 15, 12, 18, 20, 25))

# Create the ggplot object and add the scatterplot geometry
ggplot(data, aes(x = x_values, y = y_values)) +
  geom_point(color = "blue", size = 3) +
  labs(title = "Scatterplot Example", x = "X-axis", y = "Y-axis")

```

In this example, we have a sample data frame named data with two **columns x_values** and **y_values**. We created a scatterplot using **ggplot2** by specifying the data and aesthetic mappings in the **ggplot()** function and adding the scatterplot geometry with **geom_point()**. We customized the point color and size in the scatterplot. We also added a title and axis labels using **labs()**.

Remember to replace data, **x_values**, and **y_values** with your actual data frame and variable names in the **ggplot2** code. You can further customize the appearance of the plot using additional **ggplot2** functions to suit your specific visualization needs. **ggplot2** provides numerous options for creating attractive and informative scatterplots.

**Interactive graphics**

```{r}
library(ggplot2)
library(plotly)
library(gapminder)
 
p <- gapminder %>%
  filter(year==1977) %>%
  ggplot( aes(gdpPercap, lifeExp, size = pop, color=continent)) +
  geom_point() +
  scale_x_log10() +
  theme_bw()
 
ggplotly(p)
```


## Bar chart

**Bar Chart using ggplot2 in R Programming**

**ggplot2** is a popular and powerful data visualization package in R that offers a flexible and elegant way to create high-quality graphics. To create a bar chart using **ggplot2**, you need to follow a specific workflow using the **ggplot()** function along with the **geom_bar()** function.

**Step-by-Step Workflow:**

Load the necessary library: You need to load the ggplot2 library before using it.

```{r}
# Load ggplot2 library
library(ggplot2)

```

1. Prepare the data: Ensure you have the data in a data frame or tibble format with the variables you want to plot.
2. Create the ggplot object: Use the **ggplot()** function to specify the data and aesthetic mappings for the plot.
3. Add the geometry: Use the **geom_bar()** function to create the bar chart.
4. Customize the plot (optional): You can customize the appearance of the plot using various ggplot2 functions like **labs(), theme(), scale_x_discrete(), scale_y_continuous(),** etc.

**Example: Creating a Bar Chart with ggplot2**

```{r}
# Sample data
data <- data.frame(categories = c("Category 1", "Category 2", "Category 3", "Category 4"),
                   counts = c(20, 15, 30, 25))

# Create the ggplot object and add the bar chart geometry
ggplot(data, aes(x = categories, y = counts)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(title = "Bar Chart Example", x = "Categories", y = "Count")

```

In this example, we have a sample data frame named data with two columns categories and counts. We created a bar chart using ggplot2 by specifying the data and aesthetic mappings in the **ggplot()** function and adding the bar chart geometry with **geom_bar()**. We customized the fill color of the bars in the chart. We also added a title and axis labels using **labs()**.

Remember to replace data, categories, and counts with your actual data frame and variable names in the ggplot2 code. You can further customize the appearance of the plot using additional ggplot2 functions to meet your specific visualization needs. ggplot2 offers numerous options for creating visually appealing and informative bar charts.

## Boxplot 

**Boxplot using ggplot2 in R Programming**

ggplot2 is a popular data visualization package in R that offers a flexible and elegant way to create high-quality graphics. To create a boxplot using ggplot2, you need to follow a specific workflow using the **ggplot()** function along with the **geom_boxplot()** function.

**Step-by-Step Workflow:**

Load the necessary library: You need to load the **ggplot2** library before using it.

```{r}
# Load ggplot2 library
library(ggplot2)

```

1. Prepare the data: Ensure you have the data in a data frame or tibble format with the variables you want to plot.
2. Create the ggplot object: Use the **ggplot()** function to specify the data and aesthetic mappings for the plot.
3. Add the geometry: Use the **geom_boxplot()** function to create the boxplot.
4. Customize the plot (optional): You can customize the appearance of the plot using various ggplot2 functions like **labs(), theme(), scale_x_discrete(), scale_y_continuous(),** etc.

**Example: Creating a Boxplot with ggplot2**

```{r}
# Sample data
data <- data.frame(groups = rep(c("Group 1", "Group 2", "Group 3"), each = 100),
                   values = c(rnorm(100, mean = 10, sd = 2),
                              rnorm(100, mean = 12, sd = 3),
                              rnorm(100, mean = 8, sd = 2.5)))

# Create the ggplot object and add the boxplot geometry
ggplot(data, aes(x = groups, y = values)) +
  geom_boxplot(fill = "lightblue", color = "black") +
  labs(title = "Boxplot Example", x = "Groups", y = "Values")

```

In this example, we have a sample data frame named data with two columns groups and values. We created a boxplot using ggplot2 by specifying the data and aesthetic mappings in the **ggplot()** function and adding the boxplot geometry with **geom_boxplot()**. We customized the fill color and border color of the boxes in the boxplot. We also added a title and axis labels using labs().

Remember to replace data, groups, and values with your actual data frame and variable names in the ggplot2 code. You can further customize the appearance of the plot using additional ggplot2 functions to meet your specific visualization needs. ggplot2 provides a wide range of options for creating visually appealing and informative boxplots.

## Line chart

**Line Chart using ggplot2 in R Programming**

ggplot2 is a powerful data visualization package in R that provides a flexible and elegant way to create high-quality graphics. To create a line chart using ggplot2, you need to follow a specific workflow using the **ggplot()** function along with the **geom_line()** function.

**Step-by-Step Workflow:**

Load the necessary library: You need to load the **ggplot2** library before using it.

```{r}
# Load ggplot2 library
library(ggplot2)

```

1. Prepare the data: Ensure you have the data in a data frame or tibble format with the variables you want to plot.
2. Create the ggplot object: Use the **ggplot()** function to specify the data and aesthetic mappings for the plot.
3. Add the geometry: Use the **geom_line()** function to create the line chart.
4. Customize the plot (optional): You can customize the appearance of the plot using various ggplot2 functions like **labs(), theme(), scale_x_continuous(), scale_y_continuous(),** etc.

**Example: Creating a Line Chart with ggplot2**

```{r}
# Sample data
data <- data.frame(time = c(1, 2, 3, 4, 5, 6),
                   values = c(10, 15, 12, 18, 20, 25))

# Create the ggplot object and add the line chart geometry
ggplot(data, aes(x = time, y = values)) +
  geom_line(color = "blue", size = 1.5) +
  labs(title = "Line Chart Example", x = "Time", y = "Values")

```

In this example, we have a sample data frame named data with two columns time and values. We created a line chart using ggplot2 by specifying the data and aesthetic mappings in the **ggplot()** function and adding the line chart geometry with **geom_line()**. We customized the line color and size in the chart. We also added a title and axis labels using labs().

Remember to replace data, time, and values with your actual data frame and variable names in the ggplot2 code. You can further customize the appearance of the plot using additional ggplot2 functions to meet your specific visualization needs. ggplot2 provides numerous options for creating visually appealing and informative line charts.


## Lab Quiz

Topic 1,2,3 
15% of ongoing assessment
